Calendar-based graphics for visualising people's daily schedules

### 2017-07-05 main.tex
switch to jss style

---
title: "Calendar-based graphics for visualising people's daily schedules"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Di
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
fontsize: 11pt
papersize: a4
fontfamily: mathpazo
blind: false
cover: true
wpnumber: no/yr
jelcodes: C10,C14,C22
bibliography: ["references.bib", "packages.bib"]
biblio-style: authoryear-comp
abstract: "A brief summary of our ideas"
keywords: "blah, blah"
subparagraph: true
toc: false
output: 
  bookdown::pdf_document2:
    template: monashwp.tex
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    number_sections: yes
    citation_package: biblatex
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = TRUE, external = TRUE, dev = "svglite"
)
read_chunk('src/main.R')
```

```{r load}
```

# Introduction

Calendar-based visualisation proves useful in unfolding human-related activities in temporal context. For example, @van_wijk_cluster_1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It effectively contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: `ggTimeSeries` [@R-ggTimeSeries] and `ggcal` [@R-ggcal]. However, their technique is too constrained to colour-encoding graphics. We shall extend this calendar-based arrangement using linear algebra tools from a single type of heatmap to a broader range of glyphs, from univariate to multivariate cases, from one calendar format to three formats. The proposed algorithm has been developed in the `sugrrants` package [@R-sugrrants] using R [@R-base].

This work was originally motivated by studying hourly foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed across downtown Melbourne since 2009 in order to capture the pulse of city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends itself to addressing some of the issues: (1) to simultaneously exhibit multiple seasons including time of day, day of week, and day of year, (2) to visualise long historical time series data in an information-dense way at the overview level, (3) to compare and contrast multiple time series on a common scale, (4) to quickly look up the date when there are unexpected events happening without interactivity.

# Construction

@@

## Grid construction

The algorithm for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar, requires cells for days, embedded in blocks corresponding to months, organised into a grid layout for a year. Each month can be captured with 35 (5 $\times$ 7) cells, where the top left is Monday of week 1, and the bottom right is Sunday of week 5. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday-Sunday, which is determined by the year of the calendar. Months are of different length days, ranging form 28-31, and each month could extend over six weeks but the convention in these months is to wrap the last few days up to the top row of the block. The notation for creating these cells is as follows:

* $k = 1, \dots , 7$ is the day of the week that is the first day of the month
* $d = 28, 29, 30$ or $31$ representing the number of days in any month
* $(i, j)$ is the grid position where $1 \le i \le 5$ is week within the month, $1 \le j \le 7$, is day of the week. 
* $g = k, \dots,(k+d)$ indexes the day in the month, inside the 35 possible cells

@@

\begin{equation}
  \begin{aligned}
  i &= \lceil (g \mod 35) / 7\rceil, \\
  j &= g \mod 7. (\#eq:grid)
  \end{aligned}
\end{equation}

Figure \@ref(fig:month-diagram) illustrates this $(i,j)$ layout for a month where $k=5$.

```{r month-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "Illustration of the indexing layout for cells in a month."}
include_graphics("figure/month.pdf")
```

To create the layout for a full year, $(m, n)$ denotes the position of the month arranged in the plot, where $1 \le m \le M$ and $1 \le n \le N$. Between each month requires some small amount of white space, label this $b$. Figure \@ref(fig:year-diagram) illustrates this  layout.

```{r year-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "Illustration of the indexing layout for months of one year."}
include_graphics("figure/year.pdf")
```

Each cell forms a canvas on which to draw the data. Consider the canvas to have limits $[0, 1]$ horizontally and vertically. For the pedestrian sensor data, within each cell hour is plotted horizontally and count is plotted vertically. Each variable is scaled to have values between $[0,1]$, using the minimum and maximum of all the data values to be displayed. Let $h$ be the scaled hour, and $c$ the scaled count.

@@

\begin{equation}
  \begin{aligned}
  x &= i + (i - 1) \times m + (m - 1) \times b + h, \\
  y &= -j - (j - 1) \times n - (n - 1) \times b + c. (\#eq:final)
  \end{aligned}
\end{equation}

@@

The algorithm can be extended relatively easily to layout from one single month to multiple years depending on the time span to be visualised. The month-by-month arrangement can be determined by the user's choice through $M$ and $N$. If one would like to compare and contrast January, February and so on across multiple years, $M = 12$ and $N$ depending on the number of years could be setup.

@@

We only illustrate that calendar grids are laid out on the horizontal direction. The vertical direction can be enabled by swapping $i$ and $j$ in the algorithm stated above. It is useful for users in some countries where they get used to calendars of vertical organisation and other possible comparisons like Mondays across the columns rather than rows.

## Scales

@@

## Reference lines and labels

@@

Regarding the monthly calendar format, the major reference lines separate every month panel and the minor ones separate every day cell, represented by the think and thin lines respectively. The major reference lines are placed on the far left and right as well as the bottom and top of every month panel: for each $m$, the vertical lines are determined by $\min{(x)}$ and $\max{(x)}$; for each $n$, the horizontal lines are computed by $\min{(y)}$ and $\max{(y)}$. The minor reference lines are placed on the initial positions: for each $i$, the vertical line is $\min{(x)}$; for each $j$, the horizontal line is $\min{(y)}$.

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $m$ and $n$. The weekday labels are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$.

# The `frame_calendar` function

```{r args, eval = FALSE, echo = TRUE}
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h",
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

# Discussion

@@

```{r write-bib}
write_bib(c( .packages()), 'packages.bib')
```

>>>>

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Earo Wang\\Monash University \And Di Cook\\Monash Univeristy \And Rob J Hyndman\\Monash Univeristy
}
\title{Calendar-based graphics for visualising people's daily schedules}

\Plainauthor{Earo Wang, Di Cook, Rob J Hyndman}
\Plaintitle{Calendar-based graphics for visualising people's daily schedules}

\Abstract{
The abstract of the article.
}

\Keywords{calendar-based visualisation, temporal-context data}
\Plainkeywords{calendar-based visualisation, temporal-context data}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Earo Wang\\
  Monash University\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{earo.wang@monash.edu}\\

      Di Cook\\
  Monash Univeristy\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{dicook@monash.edu}\\

      Rob J Hyndman\\
  Monash Univeristy\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{rob.hyndman@monash.edu}\\

  }

% Pandoc header

\usepackage{amsmath}

\begin{document}

\section{Introduction}\label{introduction}

Calendar-based visualisation turns out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It
effectively contrasts weekdays and weekends, highlights public holiday,
and presents other known seasons like school vacations, all of which
have influence over the turn-outs in the office. The calendar-based
heatmap was implemented in a couple of R packages: \texttt{ggTimeSeries}
\citep{R-ggTimeSeries} and \texttt{ggcal} \citep{R-ggcal}. However,
their techniques are too constrained to colour-encoding graphics. We
shall extend this calendar-based arrangement using linear algebra tools
from a single type of heatmap to a broader range of glyphs, from
univariate to multivariate cases, from one calendar format to three
formats. The proposed algorithm has been developed in the
\texttt{sugrrants} package \citep{R-sugrrants} using R \citep{R-base}.

This work was originally motivated by studying hourly foot traffic in
the city of Melbourne \citep{ped}. There have been 43 sensors installed
across downtown Melbourne since 2009 in order to capture the pulse of
city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends
itself to addressing some of the issues:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  to simultaneously exhibit multiple seasons including time of day, day
  of week, and day of year
\item
  to visualise long historical time series data in an information-dense
  way at the overview level
\item
  to compare and contrast multiple time series on a common scale
\item
  to quickly look up the date when there are unexpected events happening
  without interactivity.
\end{enumerate}

\section{The function and its
construction}\label{the-function-and-its-construction}

@@

\begin{verbatim}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
\end{verbatim}

\subsection{Grid construction}\label{grid-construction}

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar, requires cells
for days, embedded in blocks corresponding to months, organised into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5. These cells provide a micro canvas on which to plot
the data. The first day of the month could be any of Monday-Sunday,
which is determined by the year of the calendar. Months are of different
length days, ranging form 28-31, and each month could extend over six
weeks but the convention in these months is to wrap the last few days up
to the top row of the block. The notation for creating these cells is as
follows:

\begin{itemize}
\tightlist
\item
  \(k = 1, \dots , 7\) is the day of the week that is the first day of
  the month
\item
  \(d = 28, 29, 30\) or \(31\) representing the number of days in any
  month
\item
  \((i, j)\) is the grid position where \(1 \le i \le 5\) is week within
  the month, \(1 \le j \le 7\), is day of the week.
\item
  \(g = k, \dots,(k+d)\) indexes the day in the month, inside the 35
  possible cells
\end{itemize}

@@

\begin{equation}
  \begin{aligned}
  i &= \lceil (g \mod 35) / 7\rceil, \\
  j &= g \mod 7. \label{eq:grid}
  \end{aligned}
\end{equation}

Figure \ref{fig:month-diagram} illustrates this \((i,j)\) layout for a
month where \(k=5\).

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=360pt,height=250pt]{figure/month}

}

\caption[Illustration of the indexing layout for cells in a month]{Illustration of the indexing layout for cells in a month.}\label{fig:month-diagram}
\end{figure}
\end{CodeChunk}

To create the layout for a full year, \((m, n)\) denotes the position of
the month arranged in the plot, where \(1 \le m \le M\) and
\(1 \le n \le N\). Between each month requires some small amount of
white space, label this \(b\). Figure \ref{fig:year-diagram} illustrates
this layout.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=360pt,height=250pt]{figure/year}

}

\caption[Illustration of the indexing layout for months of one year]{Illustration of the indexing layout for months of one year.}\label{fig:year-diagram}
\end{figure}
\end{CodeChunk}

Each cell forms a canvas on which to draw the data. Consider the canvas
to have limits \([0, 1]\) horizontally and vertically. For the
pedestrian sensor data, within each cell hour is plotted horizontally
and count is plotted vertically. Each variable is scaled to have values
between \([0,1]\), using the minimum and maximum of all the data values
to be displayed. Let \(h\) be the scaled hour, and \(c\) the scaled
count.

@@

\begin{equation}
  \begin{aligned}
  x &= i + (i - 1) \times m + (m - 1) \times b + h, \\
  y &= -j - (j - 1) \times n - (n - 1) \times b + c. \label{eq:final}
  \end{aligned}
\end{equation}

@@

The algorithm can be extended relatively easily to layout from one
single month to multiple years depending on the time span to be
visualised. The month-by-month arrangement can be determined by the
user's choice through \(M\) and \(N\). If one would like to compare and
contrast January, February and so on across multiple years, \(M = 12\)
and \(N\) depending on the number of years could be setup.

@@

We only illustrate that calendar grids are laid out on the horizontal
direction. The vertical direction can be enabled by swapping \(i\) and
\(j\) in the algorithm stated above. It is useful for users in some
countries where they get used to calendars of vertical organisation and
other possible comparisons like Mondays across the columns rather than
rows.

\subsection{Scales}\label{scales}

@@

\subsection{Reference lines and
labels}\label{reference-lines-and-labels}

@@

Regarding the monthly calendar format, the major reference lines
separate every month panel and the minor ones separate every day cell,
represented by the think and thin lines respectively. The major
reference lines are placed on the far left and right as well as the
bottom and top of every month panel: for each \(m\), the vertical lines
are determined by \(\min{(x)}\) and \(\max{(x)}\); for each \(n\), the
horizontal lines are computed by \(\min{(y)}\) and \(\max{(y)}\). The
minor reference lines are placed on the initial positions: for each
\(i\), the vertical line is \(\min{(x)}\); for each \(j\), the
horizontal line is \(\min{(y)}\).

The abbreviated month labels located on the top left are obtained
through \((\min{(x)}, \max{(y)})\) for every \(m\) and \(n\). The
weekday texts with a single letter are uniformly placed on the bottom of
the whole canvas, that is \(\min{(y)}\).

\section{Examples}\label{examples}

\section{Discussion}\label{discussion}

@@

\bibliography{packages.bib,references.bib}

\end{document}

### 2017-07-05 old_wp/main.Rmd
switch to jss style

Calendar-based visualisation proves useful in unfolding human-related activities in temporal context. For example, @van_wijk_cluster_1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It effectively contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: `ggTimeSeries` [@R-ggTimeSeries] and `ggcal` [@R-ggcal]. However, their technique is too constrained to colour-encoding graphics. We shall extend this calendar-based arrangement using linear algebra tools from a single type of heatmap to a broader range of glyphs, from univariate to multivariate cases, from one calendar format to three formats. The proposed algorithm has been developed in the `sugrrants` package [@R-sugrrants] using R [@R-base].

This work was originally motivated by studying hourly foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed across downtown Melbourne since 2009 in order to capture the pulse of city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends itself to addressing some of the issues: (1) to simultaneously exhibit multiple seasons including time of day, day of week, and day of year, (2) to visualise long historical time series data in an information-dense way at the overview level, (3) to compare and contrast multiple time series on a common scale, (4) to quickly look up the date when there are unexpected events happening without interactivity.

# Construction

@@

@@

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $m$ and $n$. The weekday labels are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$.

# The `frame_calendar` function

```{r args, eval = FALSE, echo = TRUE}
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h",
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

>>>>

Calendar-based visualisation turns out to be a useful tool in unfolding human-related activities in temporal context. For example, @van_wijk_cluster_1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It effectively contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: `ggTimeSeries` [@R-ggTimeSeries] and `ggcal` [@R-ggcal]. However, their techniques are too constrained to colour-encoding graphics. We shall extend this calendar-based arrangement using linear algebra tools from a single type of heatmap to a broader range of glyphs, from univariate to multivariate cases, from one calendar format to three formats. The proposed algorithm has been developed in the `sugrrants` package [@R-sugrrants] using R [@R-base].

This work was originally motivated by studying hourly foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed across downtown Melbourne since 2009 in order to capture the pulse of city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends itself to addressing some of the issues:

1. to simultaneously exhibit multiple seasons including time of day, day of week, and day of year 
2. to visualise long historical time series data in an information-dense way at the overview level 
3. to compare and contrast multiple time series on a common scale 
4. to quickly look up the date when there are unexpected events happening without interactivity.

# The `frame_calendar` function and its construction

@@

```{r args, eval = FALSE, echo = TRUE}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

@@

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $m$ and $n$. The weekday texts with a single letter are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$.

# Examples
